---
title: "Comparing proximity effects between conditions"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Comparing proximity effects between conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4, dpi = 90)
```

# Introduction

The `r Biocpkg("PACE")` manual fits a single section and asks how expression
changes with proximity to each neighbouring cell type. This vignette asks the
next question: **does that proximity effect differ between two groups of
patients?**

Answering it needs a cohort rather than a section, and it adds one component to
the model. Alongside the spatial cell state block, which is the proximity effect
common to everyone, PACE estimates a **responder spatial state** block: the part
of the proximity effect that differs between conditions. Read the two together,
a gene can respond to a neighbour in both arms equally (spatial, no responder
signal), or respond differently depending on outcome (responder signal).

# The melanoma subset

The package ships a subset of a CosMx melanoma cohort of patients profiled after
immune checkpoint blockade, contrasting progressive disease (PD) against
non-progressive disease (non-PD, pooling stable, partial and complete
responders).

```{r load, message = FALSE}
library(PACE)
library(SpatialExperiment)
library(ggplot2)
library(dplyr)

spe <- readRDS(system.file("extdata", "mel_cosmx_subset.rds", package = "PACE"))
spe
table(spe$cellType)
table(unique(data.frame(image = spe$image, arm = spe$Responder))$arm)
```

One section per patient, so the condition contrast is between patients, not
within a section. Each section here is a contiguous spatial crop of the
published one; the derivation is in `inst/scripts/make-mel-cosmx-subset.R`.

> **Provenance.** These data are a modified subset of the deposit accompanying
> Dong, Su, Kluger, Fan and Kluger, *SIMVI disentangles intrinsic and
> spatial-induced cellular states in spatial omics data*
> ([doi:10.5281/zenodo.14708000](https://doi.org/10.5281/zenodo.14708000)),
> used here under CC-BY-4.0.

> **Read this before drawing conclusions.** The subset is 8,454 of the cohort's
> 56,274 cells, sized so the package stays small enough to distribute. It is an
> illustration of the workflow, not an analysis. The section on statistical
> power below shows what that costs, with numbers.

# Fitting with a condition

The call is the single-section one plus three arguments:

- `condition_col` names the column holding the contrast, and switches on the
  responder spatial state block.
- `kernel_per_image = TRUE` builds neighbourhoods within each section, so no
  cell is ever a neighbour of a cell from another patient.
- `image_re = "intercept"` gives each section a random intercept, so
  patient-level differences in overall expression are absorbed rather than
  being attributed to the condition.

```{r fit}
fit <- paceFit(spe,
               celltype_col     = "cellType",
               condition_col    = "Responder",
               image_col        = "image",
               kernel_per_image = TRUE,
               image_re         = "intercept",
               dispersion       = "nb1",
               verbose          = FALSE)
fit
```

The interaction term is named from the non-reference level of `condition_col`:

```{r respterm}
fit@params$resp_term
```

so `ResponderPD:Tumour` is the extra change in expression per unit of tumour
proximity in the PD arm, on top of the `Tumour` effect shared by both arms.

# The responder block in the decomposition

`varianceDecomposition()` now carries a `Responder spatial %` column beside the
`Spatial %` one:

```{r decomp}
varianceDecomposition(fit) |>
  group_by(focal) |>
  summarise(`Spatial %`           = round(median(`Spatial %`), 4),
            `Responder spatial %` = round(median(`Responder spatial %`), 4),
            `Spillover %`         = round(median(`Spillover %`), 2),
            .groups = "drop")
```

The responder block is far smaller than the shared spatial block, which is the
expected shape: most of how a cell responds to its neighbours is common to both
arms, and only a thin slice of it depends on outcome. Contamination is larger
than either, and largest in the sparsely distributed immune populations.

# Which pairs carry the condition difference

`pairVariance()` and `plotPairHeatmap()` both take `block = "responder"`, which
attributes the responder block across neighbours instead of the shared spatial
one:

```{r pairs-responder, fig.width = 6, fig.height = 4.6}
pairVariance(fit, block = "responder") |>
  arrange(desc(val)) |>
  head(5)

plotPairHeatmap(fit, block = "responder")
```

Macrophages next to tumour cells is among the strongest pairs here, which is the
relationship the accompanying paper reports for the full cohort. Compare it with
the shared spatial block on the same fit:

```{r pairs-spatial, fig.width = 6, fig.height = 4.6}
plotPairHeatmap(fit, block = "spatial")
```

# Statistical power, and what this subset cannot show

The condition contrast is between patients, and the outcome-dependent part of
the spatial signal is sparse. Both facts mean it needs the whole cohort. In this
subset, no gene reaches the usual significance threshold:

```{r calls}
ns <- neighbourSlopes(fit)
c(rows = nrow(ns), calls_lfsr_under_0.05 = sum(ns$lfsr < 0.05, na.rm = TRUE))
```

`topDrivers()` scores every pair and labels each one, and its own verdict for
all thirty here is the right one:

```{r drivers}
table(vapply(topDrivers(fit), function(x) x$status, character(1)))
```

## What the full cohort gives

Rather than assert the difference, the package ships the macrophage slopes from
the **full** cohort, all 56,274 cells across these same 26 patients, so the two
can be put side by side. It is a fitted result, not something this vignette
computes; the `provenance` attribute records how:

```{r fullcohort}
full <- readRDS(system.file("extdata", "mel_full_cohort_macrophage_slopes.rds",
                            package = "PACE"))
nrow(full)
subset(full, gene == "SPP1" & neighbour == "Tumour" & term == "ResponderPD:Tumour")
```

That is the result the paper reports: in macrophages, the response of *SPP1* to
tumour proximity differs between arms, more negative in progressive disease. Now
the same row from the subset fitted above:

```{r subsetrow}
subset(ns, gene == "SPP1" & focal == "Macrophage" &
           neighbour == "Tumour" & term == "ResponderPD:Tumour")
```

The estimate has been shrunk to exactly zero. This is worth being precise about,
because it is not a threshold that could be relaxed: 32,445 of the 33,372
responder terms are shrunk to zero here, mash having concluded that the block
carries no signal at this sample size. No `lfsr` cut recovers *SPP1*, because
there is no estimate left to recover. The unshrunk estimate does keep the right
sign, but it is smaller than its own standard error.

Cropping each section to 15% removes the effect entirely; it only begins to
reappear at around 60% of the cohort, and even there it is less than half its
full size. Crops targeted at the macrophages do recover it, but only by keeping
most of the tissue, because the macrophages are spread through all of it.

So the pair ranking above is worth reading as a direction, and the significance
column as a warning. For the published result, fit the full deposit.

The practical lesson generalises beyond this dataset: a condition analysis is
powered by **patients**, not by cells. Adding cells from the patients you
already have sharpens each patient's estimate, but the contrast is still a
comparison of two small groups.

# Session info

```{r sessioninfo}
sessionInfo()
```
