---
title: "Comparing proximity effects between conditions"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Comparing proximity effects between conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4, dpi = 90)
```

# Introduction

The `r Biocpkg("PACE")` manual fits a single section and asks how expression
changes with proximity to each neighbouring cell type. This vignette asks the
next question: **does that proximity effect differ between two groups of
patients?**

Answering it needs a cohort rather than a section, and it adds one component to
the model. Alongside the spatial cell state block, which is the proximity effect
common to everyone, PACE estimates a **responder spatial state** block: the part
of the proximity effect that differs between conditions. Read the two together,
a gene can respond to a neighbour in both arms equally (spatial, no responder
signal), or respond differently depending on outcome (responder signal).

# The melanoma subset

The package ships a subset of a CosMx melanoma cohort of patients profiled after
immune checkpoint blockade, contrasting progressive disease (PD) against
non-progressive disease (non-PD, pooling stable, partial and complete
responders).

```{r load, message = FALSE}
library(PACE)
library(SpatialExperiment)
library(ggplot2)
library(dplyr)

spe <- readRDS(system.file("extdata", "mel_cosmx_subset.rds", package = "PACE"))
spe
table(spe$cellType)
table(unique(data.frame(image = spe$image, arm = spe$Responder))$arm)
```

One section per patient, so the condition contrast is between patients rather
than within a section. That is what makes a cohort necessary: the effect is
estimated across patients, and no number of extra cells from the same patients
substitutes for it.

Every cell and every patient of the published cohort is here. What has been
reduced is the panel: 180 of the 927 genes, the most widely detected ones, kept
so the data are small enough to ship. Cells and neighbourhoods are untouched,
which is what the model reads; the derivation is in
`inst/scripts/make-mel-cosmx-subset.R`.

> **Provenance.** These data are a subset of the deposit accompanying Dong, Su,
> Kluger, Fan and Kluger, *SIMVI disentangles intrinsic and spatial-induced
> cellular states in spatial omics data*
> ([doi:10.5281/zenodo.14708000](https://doi.org/10.5281/zenodo.14708000)),
> used here under CC-BY-4.0.

# Fitting with a condition

The call is the single-section one plus three arguments:

- `condition_col` names the column holding the contrast, and switches on the
  responder spatial state block.
- `kernel_per_image = TRUE` builds neighbourhoods within each section, so no
  cell is ever a neighbour of a cell from another patient.
- `image_re = "intercept"` gives each section a random intercept, so
  patient-level differences in overall expression are absorbed rather than
  being attributed to the condition.

```{r fit}
fit <- paceFit(spe,
               celltype_col     = "cellType",
               condition_col    = "Responder",
               image_col        = "image",
               kernel_per_image = TRUE,
               image_re         = "intercept",
               dispersion       = "nb1",
               verbose          = FALSE)
fit
```

The interaction term is named from the non-reference level of `condition_col`:

```{r respterm}
fit@params$resp_term
```

so `ResponderPD:Tumour` is the extra change in expression per unit of tumour
proximity in the PD arm, on top of the `Tumour` effect shared by both arms.

# The responder block in the decomposition

`varianceDecomposition()` now carries a `Responder spatial %` column beside the
`Spatial %` one:

```{r decomp}
varianceDecomposition(fit) |>
  group_by(focal) |>
  summarise(`Spatial %`           = round(median(`Spatial %`), 4),
            `Responder spatial %` = round(median(`Responder spatial %`), 4),
            `Spillover %`         = round(median(`Spillover %`), 2),
            .groups = "drop")
```

The responder block is far smaller than the shared spatial block, which is the
expected shape: most of how a cell responds to its neighbours is common to both
arms, and only a thin slice of it depends on outcome. Contamination is larger
than either, and largest in the sparsely distributed immune populations.

# Which pairs carry the condition difference

`pairVariance()` and `plotPairHeatmap()` both take `block = "responder"`, which
attributes the responder block across neighbours instead of the shared spatial
one:

```{r pairs-responder, fig.width = 6, fig.height = 4.6}
pairVariance(fit, block = "responder") |>
  arrange(desc(val)) |>
  head(5)

plotPairHeatmap(fit, block = "responder")
```

Fibroblasts next to endothelium carries the largest share, with tumour and
macrophage pairs behind it; the responder signal is concentrated in stromal and
myeloid populations rather than spread evenly. Compare it with the shared
spatial block on the same fit, where the ordering is different:

```{r pairs-spatial, fig.width = 6, fig.height = 4.6}
plotPairHeatmap(fit, block = "spatial")
```

# The genes that carry it

Pair shares say where to look; `neighbourSlopes()` says which genes. The
`ResponderPD:` terms are the ones that differ between arms:

```{r calls}
ns <- neighbourSlopes(fit)
resp <- subset(ns, grepl("^ResponderPD:", term) & lfsr < 0.05)
nrow(resp)

resp[order(resp$lfsr), c("gene", "focal", "neighbour", "estimate_shrunk", "lfsr")] |>
  head(8)
```

Among the strongest is *SPP1* in macrophages next to tumour cells:

```{r spp1}
subset(ns, gene == "SPP1" & focal == "Macrophage" & neighbour == "Tumour")
```

The `Tumour` row is the response shared by both arms; the `ResponderPD:Tumour`
row is how much that response differs in progressive disease. It is negative,
so macrophage *SPP1* rises less steeply with tumour proximity in patients whose
disease progressed. This is the melanoma result reported in the accompanying
paper, recovered here from the shipped data.

`plotProximity()` shows the counts behind it. The default bin breaks are set for
sparser tissue; this section is dense enough that a macrophage has a median of 22
tumour cells within 30 um, so the bins are widened to spread the cells out
rather than piling 83% of them into the last one:

```{r proximity, fig.width = 6.5, fig.height = 4.2}
plotProximity(fit, spe, "SPP1", "Macrophage", "Tumour",
              breaks = c(0, 10, 20, 30, 40, Inf))
```

That is the response shared by both arms, rising steeply with tumour proximity.
The condition term above says the rise is shallower in progressive disease.

# A note on the reduced panel

Fitting 180 of the 927 genes leaves the estimates alone but not the shrinkage.
mash calibrates against the genes it is given, so a narrower panel shifts the
local false sign rates: this fit reports `r nrow(resp)` responder calls where the
full panel reports 46. The *SPP1* estimate is essentially unchanged, −0.0663
against −0.0668 on the full panel, but treat the significance column here as
belonging to this panel rather than to the published analysis, and fit the full
deposit if you need to reproduce the paper exactly.

The panel cannot be cut much further. Below roughly 150 genes some focal cell
type runs out of genes to decompose and `paceDecompose()` fails outright.

# Session info

```{r sessioninfo}
sessionInfo()
```
