---
title: "Quantifying proximity effects with PACE"
author: "Elijah Willie"
package: PACE
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Quantifying proximity effects with PACE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`r Biocpkg("PACE")` (Proximity-Associated Changes in Expression) quantifies how a
cell's gene expression changes with proximity to specific neighbouring cell types
in imaging-based spatial transcriptomics. It fits hierarchical negative binomial
mixed models with partial pooling across cell types, corrects for transcript
contamination between adjacent cells, and decomposes expression variance into
cell-type identity, spatial cell state, contamination, and residual components.

Input follows `r Biocpkg("SpatialExperiment")` conventions.

# Installation

```{r install, eval = FALSE}
if (!require("BiocManager")) install.packages("BiocManager")
BiocManager::install("PACE")
```

# A worked example on synthetic data

We build a small synthetic `SpatialExperiment` with three cell types.

```{r toy, message = FALSE}
library(PACE)
library(SpatialExperiment)

set.seed(1)
n <- 600
types <- c("A", "B", "C")
ct <- sample(types, n, replace = TRUE)
coords <- cbind(x = runif(n, 0, 500), y = runif(n, 0, 500))
counts <- do.call(rbind, lapply(types, function(t) {
  lam <- ifelse(ct == t, 8, 1)
  t(vapply(seq_len(12), function(i) rpois(n, lam), numeric(n)))
}))
rownames(counts) <- paste0("g", seq_len(nrow(counts)))
colnames(counts) <- paste0("c", seq_len(n))

spe <- SpatialExperiment(
  assays        = list(counts = counts),
  colData       = DataFrame(cellType = ct),
  spatialCoords = coords)
spe
```

# Fitting the model

```{r fit}
fit <- paceFit(spe, celltype_col = "cellType",
               contamination = "none", n_iter = 5, verbose = FALSE)
fit
```

# Reading the results

The shrunken per-(gene, focal, neighbour) proximity slopes with their local false
sign rate:

```{r slopes}
head(neighbourSlopes(fit))
```

The per-gene variance decomposition:

```{r decomp}
head(varianceDecomposition(fit))
```

The per-pair driver-score tables:

```{r drivers}
names(topDrivers(fit))
```

# Multi-sample, condition-stratified cohorts

For a cohort of multiple images or patients with a treatment/response condition,
supply the grouping and condition columns and enable the per-image kernels and
the responder random-effect block:

```{r multi, eval = FALSE}
fit <- paceFit(spe,
               celltype_col     = "celltype",
               image_col        = "sampleID",
               condition_col     = "response",
               kernel_per_image = TRUE,
               image_re         = "condition_slopes")
```

# Session info

```{r sessioninfo}
sessionInfo()
```
