---
title: "PACE package manual"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{PACE package manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4, dpi = 90)
```

# Installation

```{r install, eval = FALSE}
if (!require("BiocManager")) install.packages("BiocManager")
BiocManager::install("remotes")
remotes::install_github("ecool50/PACE")
```

# Introduction

`r Biocpkg("PACE")` (Proximity-Associated Changes in Expression) is a framework
for imaging-based spatial transcriptomics (such as Xenium and CosMx) that
quantifies how a cell's gene expression changes with proximity to *specific*
neighbouring cell types. Rather than asking only which genes are spatially
variable, PACE asks a directed question: when a macrophage sits next to tumour
cells, which of its genes go up or down, and by how much?

PACE fits a hierarchical negative binomial mixed model with partial pooling
across cell types, so that each focal-neighbour relationship borrows strength
from the others and noisy pairs are regularised towards the consensus. A per-cell
contamination term absorbs the short-range ambient signal that leaks between
adjacent cells through segmentation error, so the estimated neighbour effect
reflects biology rather than transcript misassignment.

# Overview of the PACE workflow

A PACE analysis has three stages, all driven by the single call `paceFit()`:

1. **Neighbourhood construction.** For every cell, a Gaussian biological kernel
   summarises the abundance of each neighbouring cell type, and a short-range
   exponential technical kernel captures the ambient contamination field.
2. **Model fitting.** A streaming penalised quasi-likelihood fit estimates, for
   every gene and every focal cell type, a partially pooled slope on each
   neighbour type's abundance, jointly with a per-cell contamination loading and
   gene-wise overdispersions. The neighbour slopes are then stabilised by
   multivariate adaptive shrinkage.
3. **Interpretation.** The fit is read out three ways: shrunken neighbour slopes
   with a local false sign rate (`neighbourSlopes()`), a per-gene variance
   decomposition (`varianceDecomposition()`), and per-pair driver-score tables
   ranking the genes that mediate each relationship (`topDrivers()`).

# The breast cancer dataset

The package ships a small worked example: a spatial crop of a human breast cancer
Xenium section (10x Genomics; segmented with BIDCell), centred on a
tumour-macrophage interface. It contains `r "7,898"` cells across eight cell
types (B cell, dendritic cell, endothelial, macrophage, myoepithelial, stromal,
T cell, and tumour) and a 278-gene panel, packaged as a
`r Biocpkg("SpatialExperiment")`. This region reproduces the macrophage
reprogramming that PACE identifies on the full section.

```{r load, message = FALSE}
library(PACE)
library(SpatialExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)

spe <- readRDS(system.file("extdata", "bc_xenium_subset.rds", package = "PACE"))
spe
table(spe$cellType)
```

# Fitting the model

`paceFit()` reads the counts, spatial coordinates, and cell-type labels from the
`SpatialExperiment` and runs the whole pipeline. On this crop it takes about half
a minute.

```{r fit}
fit <- paceFit(spe,
               celltype_col  = "cellType",
               contamination = "percell_hc",   # per-cell contamination correction
               dispersion    = "nb1",
               verbose       = FALSE)
fit
```

# Variance decomposition

`plotDecomposition()` shows, for each focal cell type, the share of expression
variance attributable to cell-type identity, spatial cell state (proximity
effects), and contamination (spillover), pooled over genes. The left panel is the
full stacked bar; the right zooms in on the two small spatial components.

```{r decomp, fig.width = 11, fig.height = 4.5}
plotDecomposition(fit)
```

Beyond the dominant cell-type identity component, macrophages, stromal and
myoepithelial cells carry the largest spatial cell-state contributions. The
underlying per-gene table is available with `varianceDecomposition(fit)`.

# Pairwise spatial interactions

`plotPairHeatmap()` gives the percentage of each focal type's total variance
contributed by spatial interaction with each neighbour, computed as the focal's
spatial share split across neighbours by a normalised Pratt attribution.
**Tumour as a neighbour** drives the strongest spatial signal across the
microenvironment.

```{r pairs, fig.width = 6.5, fig.height = 5}
plotPairHeatmap(fit)
```

# Macrophage–tumour drivers

`plotDrivers()` ranks the genes mediating a relationship by a driver score (MCSD,
combining the shrunken slope, its cell-type specificity, and its expression
level), alongside their per-gene single-frame decomposition. For the
macrophage-tumour pair, the top drivers are the tissue-resident marker **MRC1**
(CD206), reduced near tumour, and the lipid-associated marker **APOC1**, elevated
near tumour.

```{r drivers, fig.width = 10, fig.height = 4.5}
plotDrivers(fit, "Macrophage", "Tumour")
```

The shrunken slopes confirm the opposing directions:

```{r flagships}
neighbourSlopes(fit) |>
  filter(focal == "Macrophage", neighbour == "Tumour",
         gene %in% c("MRC1", "APOC1")) |>
  select(gene, estimate_shrunk, lfsr)
```

# Visualising the proximity effect

`plotProximity()` bins each macrophage by its kernel-weighted tumour-neighbour
density and shows each gene's log1p CP10k expression per bin. Because these genes
are zero-inflated (the box median sits at zero in most bins), the per-bin mean is
overlaid as a diamond and connecting line to make the trend explicit, alongside
the percentage of macrophages expressing the gene and the fitted
expression-versus-density slope. MRC1 falls and APOC1 rises with tumour proximity.

```{r gradient, fig.width = 11, fig.height = 4.6}
plotProximity(fit, spe, c("MRC1", "APOC1"), "Macrophage", "Tumour")
```

# Biological interpretation

On this section PACE recovers a coordinated shift in macrophage state at the
tumour interface: the tissue-resident marker MRC1 (CD206) is reduced and the
lipid-associated, immunosuppressive marker APOC1 is elevated in macrophages near
tumour cells. Because PACE separates this proximity effect from the transcript
contamination that would otherwise inflate tumour-marker signal in the
macrophages, the drivers it promotes are genuine macrophage genes rather than
tumour genes bleeding across cell boundaries.

# Session info

```{r sessioninfo}
sessionInfo()
```
