---
title: "PACE package manual"
author: "Elijah Willie"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{PACE package manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 6, fig.height = 4, dpi = 90)
```

# Installation

```{r install, eval = FALSE}
if (!require("BiocManager")) install.packages("BiocManager")
BiocManager::install("remotes")
remotes::install_github("ecool50/PACE")
```

# Introduction

`r Biocpkg("PACE")` (Proximity-Associated Changes in Expression) is a framework
for imaging-based spatial transcriptomics (such as Xenium and CosMx) that
quantifies how a cell's gene expression changes with proximity to *specific*
neighbouring cell types. Rather than asking only which genes are spatially
variable, PACE asks a directed question: when a macrophage sits next to tumour
cells, which of its genes go up or down, and by how much?

PACE fits a hierarchical negative binomial mixed model with partial pooling
across cell types, so that each focal-neighbour relationship borrows strength
from the others and noisy pairs are regularised towards the consensus. A per-cell
contamination term absorbs the short-range ambient signal that leaks between
adjacent cells through segmentation error, so the estimated neighbour effect
reflects biology rather than transcript misassignment.

# Overview of the PACE workflow

A PACE analysis has three stages, all driven by the single call `paceFit()`:

1. **Neighbourhood construction.** For every cell, a Gaussian biological kernel
   summarises the abundance of each neighbouring cell type, and a short-range
   exponential technical kernel captures the ambient contamination field.
2. **Model fitting.** A streaming penalised quasi-likelihood fit estimates, for
   every gene and every focal cell type, a partially pooled slope on each
   neighbour type's abundance, jointly with a per-cell contamination loading and
   gene-wise overdispersions. The neighbour slopes are then stabilised by
   multivariate adaptive shrinkage.
3. **Interpretation.** The fit is read out three ways: shrunken neighbour slopes
   with a local false sign rate (`neighbourSlopes()`), a per-gene variance
   decomposition (`varianceDecomposition()`), and per-pair driver-score tables
   ranking the genes that mediate each relationship (`topDrivers()`).

# The breast cancer dataset

The package ships a small worked example: a spatial crop of a human breast cancer
Xenium section (10x Genomics; segmented with BIDCell), centred on a
tumour-macrophage interface. It contains `r "7,898"` cells across eight cell
types (B cell, dendritic cell, endothelial, macrophage, myoepithelial, stromal,
T cell, and tumour) and a 278-gene panel, packaged as a
`r Biocpkg("SpatialExperiment")`. This region reproduces the macrophage
reprogramming that PACE identifies on the full section.

```{r load, message = FALSE}
library(PACE)
library(SpatialExperiment)

spe <- readRDS(system.file("extdata", "bc_xenium_subset.rds", package = "PACE"))
spe
table(spe$cellType)
```

# Fitting the model

`paceFit()` reads the counts, spatial coordinates, and cell-type labels from the
`SpatialExperiment` and runs the whole pipeline. On this crop it takes about half
a minute.

```{r fit}
fit <- paceFit(spe,
               celltype_col  = "cellType",
               contamination = "percell_hc",   # per-cell contamination correction
               dispersion    = "nb1",
               verbose       = FALSE)
fit
```

# Variance decomposition

`varianceDecomposition()` returns, for each focal cell type and gene, the share
of expression variance attributable to cell-type identity, spatial cell state
(proximity effects), contamination, and residual. Averaging over genes gives a
per-cell-type summary.

```{r decomp, fig.height = 3.4}
library(ggplot2)
library(dplyr)
library(tidyr)

vd <- varianceDecomposition(fit)
blocks <- c("Spatial %", "Spillover %", "Residual %")
agg <- vd |>
  group_by(focal) |>
  summarise(across(all_of(blocks), \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  pivot_longer(-focal, names_to = "component", values_to = "pct")

ggplot(agg, aes(focal, pct, fill = component)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "mean % of expression variance",
       title = "Non-identity variance components by focal cell type") +
  theme_minimal(base_size = 11)
```

Beyond the dominant cell-type-identity component (omitted here for scale),
macrophages, stromal and myoepithelial cells carry the largest spatial cell-state
contributions.

# Pairwise spatial interactions

Which focal-neighbour pairs are most strongly coupled? Counting the significant
genes (local false sign rate < 0.05) per pair shows that **tumour as a
neighbour** drives the strongest spatial signal across the microenvironment.

```{r pairs, fig.height = 4}
ns <- neighbourSlopes(fit)
pair <- ns |>
  filter(focal != neighbour) |>
  group_by(focal, neighbour) |>
  summarise(n_sig = sum(lfsr < 0.05, na.rm = TRUE), .groups = "drop")

ggplot(pair, aes(neighbour, focal, fill = n_sig)) +
  geom_tile() +
  scale_fill_viridis_c(option = "inferno") +
  labs(x = "neighbour cell type", y = "focal cell type",
       fill = "significant\ngenes",
       title = "Spatial coupling: significant genes per pair") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Macrophage–tumour drivers

`topDrivers()` ranks the genes mediating each relationship by a driver score that
combines the shrunken slope, its cell-type specificity, and its expression level.
For the macrophage-tumour pair, the top drivers are the tissue-resident marker
**MRC1** (CD206), reduced near tumour, and the lipid-associated marker **APOC1**,
elevated near tumour.

```{r drivers}
drv <- topDrivers(fit)[["Macrophage_Tumour"]]$scores
head(drv[, c("rank", "gene", "b_clean", "spec", "lfsr", "MCSD")], 8)
```

The shrunken slopes confirm the opposing directions:

```{r flagships}
ns |>
  filter(focal == "Macrophage", neighbour == "Tumour",
         gene %in% c("MRC1", "APOC1")) |>
  select(gene, estimate_shrunk, lfsr)
```

# Visualising the proximity effect

We can see the gradient directly. For each macrophage we count its tumour
neighbours within 30 um and plot MRC1 and APOC1 expression against it: MRC1 falls
and APOC1 rises with tumour proximity.

```{r gradient, fig.height = 3.6}
library(Matrix)

counts <- assay(spe, "counts")
cp10k  <- log1p(t(t(counts) / pmax(colSums(counts), 1) * 1e4))
coords <- spatialCoords(spe)
ct     <- spe$cellType

nn <- dbscan::frNN(coords, eps = 30)
tum_nbr <- vapply(nn$id, function(ii) sum(ct[ii] == "Tumour"), integer(1))

mac <- which(ct == "Macrophage")
grad <- data.frame(
  tumour_neighbours = cut(tum_nbr[mac], c(-1, 0, 2, 5, Inf),
                          labels = c("0", "1-2", "3-5", "6+")),
  MRC1  = cp10k["MRC1",  mac],
  APOC1 = cp10k["APOC1", mac]) |>
  pivot_longer(c(MRC1, APOC1), names_to = "gene", values_to = "expression")

ggplot(grad, aes(tumour_neighbours, expression, fill = gene)) +
  geom_boxplot(outlier.size = 0.3) +
  facet_wrap(~gene, scales = "free_y") +
  labs(x = "tumour neighbours (within 30 um)",
       y = "log1p CP10k expression",
       title = "Macrophage MRC1 and APOC1 vs tumour proximity") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

A spatial view of the same region, with macrophages coloured by APOC1 over the
tumour background:

```{r map, fig.height = 4.2}
plt <- data.frame(x = coords[, 1], y = coords[, 2], cellType = ct,
                  APOC1 = cp10k["APOC1", ])
ggplot() +
  geom_point(data = subset(plt, cellType == "Tumour"),
             aes(x, y), colour = "grey85", size = 0.3) +
  geom_point(data = subset(plt, cellType == "Macrophage"),
             aes(x, y, colour = APOC1), size = 0.7) +
  scale_colour_viridis_c(option = "magma") +
  coord_equal() +
  labs(title = "Macrophage APOC1 over tumour (grey)", x = "x (um)", y = "y (um)") +
  theme_minimal(base_size = 10)
```

# Biological interpretation

On this section PACE recovers a coordinated shift in macrophage state at the
tumour interface: the tissue-resident marker MRC1 (CD206) is reduced and the
lipid-associated, immunosuppressive marker APOC1 is elevated in macrophages near
tumour cells. Because PACE separates this proximity effect from the transcript
contamination that would otherwise inflate tumour-marker signal in the
macrophages, the drivers it promotes are genuine macrophage genes rather than
tumour genes bleeding across cell boundaries.

# Session info

```{r sessioninfo}
sessionInfo()
```
